A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-01 paper.tex
Deleting paragraph headings

\paragraph*{Properties of the coalescent with recombination and related models}
An extensive body of work has developed in
studying the coalescent with recombination 
% Can we add a couple of the most well known CwR examples here
and other related 
graph-valued stochastic processes from a mathematical perspective.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection instead of recombination.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode
desired patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
% Could mention the Ancestral Gene Transfer graph here too, within
% something like "Similarly the Ancestral Gene Transfer Graph [cite]
% and [other ancestral graphs, if they exist?] model genealogical
% processes in bacteria as a graph.

\paragraph*{Inference of evolutionary parameters and genealogies}
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
Several breakthroughs in this direction were achieved through
formulating simplified but more tractable approximations to the full 
model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
paul2011accurate,schiffels2014inferring}.
The related problem of \emph{sampling} genealogies compatible with a given 
dataset under the coalescent with recombination also proved notoriously difficult
computationally; progress in explicitly inferring genealogies at scale 
has similarly been achieved through resorting to principled
approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
or moving away from the coalescent with recombination altogether and seeking
to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.

\paragraph*{Optimisation problems for phylogenetic networks}
There has also been substantial interest in formulating and answering 
fundamental questions about properties
of the ARG as a discrete graph structure, abstracting away from the
generating process and focussing on the ARG topology without considering 
branch lengths. 
The first prominent problem was calculating (lower bounds on) the minimum number of
recombinations required to reconstruct a valid genealogy for a given
sample~\citep{myers2003bounds}, and constructing the corresponding 
minimal (parsimonious)
ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
focus has been on algorithmic and combinatorial results, that largely do not
have direct relevance to the inference problems described above.

@@

\paragraph*{} The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the 
stochastic process generating genealogies in the presence of 
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent}, 
as well as, more commonly, the concrete realisations of the 
ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

>>>>

An extensive body of work has developed in
studying the coalescent with recombination 
% Can we add a couple of the most well known CwR examples here
and other related 
graph-valued stochastic processes from a mathematical perspective.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection instead of recombination.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode
desired patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
% Could mention the Ancestral Gene Transfer graph here too, within
% something like "Similarly the Ancestral Gene Transfer Graph [cite]
% and [other ancestral graphs, if they exist?] model genealogical
% processes in bacteria as a graph.

Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
Several breakthroughs in this direction were achieved through
formulating simplified but more tractable approximations to the full 
model~\citep{mcvean2005approximating,marjoram2006fast,li2011inference,
paul2011accurate,schiffels2014inferring}.
The related problem of \emph{sampling} genealogies compatible with a given 
dataset under the coalescent with recombination also proved notoriously difficult
computationally; progress in explicitly inferring genealogies at scale 
has similarly been achieved through resorting to principled
approximations~\citep{rasmussen2014genome,mahmoudi2022bayesian},
or moving away from the coalescent with recombination altogether and seeking
to infer a single plausible ARG~\citep[e.g.][]{speidel2019method} or ARG
topology~\citep[e.g.][]{minichiello2006mapping,kelleher2019inferring}.

There has also been substantial interest in formulating and answering 
fundamental questions about properties
of the ARG as a discrete graph structure, abstracting away from the
generating process and focussing on the ARG topology without considering 
branch lengths. 
The first prominent problem was calculating (lower bounds on) the minimum number of
recombinations required to reconstruct a valid genealogy for a given
sample~\citep{myers2003bounds}, and constructing the corresponding 
minimal (parsimonious)
ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
focus has been on algorithmic and combinatorial results, that largely do not
have direct relevance to the inference problems described above.

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the 
stochastic process generating genealogies in the presence of 
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent}, 
as well as, more commonly, the concrete realisations of the 
ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.
